Learning by Teaching a Robot:\\The Case of Handwriting

### 2015-12-15 main.tex
Added quote of Vincent's mother + anonymized children name + plenty of little changes

Thomas is five and an half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
workaround his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{Children's names have been changed.}

While Thomas is lively and always quick at shifting his attention from one
activity to another, Vincent is shy and poised. Two very different children,
facing however the same difficulty to write in a legible manner. And, hidden
beyond this impaired skill, psycho-social difficulties also arise: they
underperform at school, Thomas has to go for follow-up visits every week,
they both live under the label ``requires special care''. This is a source of
anxiety for the children and for their parents alike.

@@

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks, \cite{Hoy2011}), essentially consisting in
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving his/her handwriting skills and,
consequently, improving his/her performance at school.

We present in this article a new take on this educative challenge, as a
remediation procedure that involve a ``bad writer'' robot that is \emph{taught}
by the child: by building on the \emph{learning by teaching} paradigm, not only
the child practises handwriting but, as (s)he takes on the role of the teacher,
(s)he also positively reinforce his/her self-esteem and motivation: his/her
social role shifts from the ``underperformer'' to ``the one who knows and
teaches''.

@@

We also investigate here a particular role for a robot in the
education of handwriting: not only is the robot actively performing the activity
by drawing letters, but it does so in a way that engages the child in a very
specific social role. The child is the teacher in this relationship and the
robot is the learner: the child must engage in a (meta-) cognitive relationship
with the robot to try to understand why the robot fails and how to help it best.
Here, the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphising, and
cognitively engage the child into the learning activity.

Besides the commitment of the child into the interaction build on the
``protégé effect'': the teacher feels responsible for his student, commits
to the student's success and possibly experiences student's failure as his own
failure to teach. Teachable computer-based agents have previously been used to
encourage this ``protégé effect'', wherein students invest more effort into
learning when it is for a teachable agent than for themselves~\cite{Chase2009}.
We rely on this cognitive mechanism to reinforce the child's commitment into the
robot-mediated handwriting activity.

@@

In our case, the role of agency is stronger: it induces meta-cognition (``I am
interacting with an agent, so I need to reflect on how to best teach him'')
which is beneficial for the learning process; it also induces a protégé
effect (``I want my robot-agent to succeed!'') which support the commitment of
child into the interaction, also for longer period of time (we report in this
article upon two case studies where children were indeed interacting with the
same robot for several hours, spanned over several weeks).

@@

The following sections of the paper go into details. We first provides a
brief overview of the robotic system and the interaction it induces. We then
give the details of machine-learning techniques that allow the robot to learn
from the children, followed by the implementation on a {\sc nao} robot.

We then present and report on the four school experiments and the two case
studies that we conducted over the last two years. While the focus of the school
experiments was mostly technical validation and data acquisition, the two case
studies that lasted each about 4 weeks, gave us initial insights on the
relevance and effectiveness of our approach.

@@

Figure~\ref{experimental_setup} illustrates our general experimental setup: a
face-to-face child-robot interaction with an (autonomous) Aldebran's {\sc nao}
robot.

A tactile tablet (with a custom application) is used for both the robot and the
child to write: during a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and push the tablet
towards the robot, the robot writes on the tablet by gesturing the writing (but
without actually physically touching the tablet), the child then pull back the
tablet, corrects the robot's attempt by writing him/herself on top or next to
the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
demonstration to the robot by pressing a small button on the tablet. The robot
``learns'' from this demonstration and tries again.

Since the child is assumed to take on the role of the teacher, we had to ensure
(s)he would be able to manage by him/herself the turn-taking and the overall
progression of the activity (moving to the next letter or word). In our design,
the turn-taking relies on the robot prompting for feedback once it is done with
its writing (simple sentences like ``What do you think?''), and pressing on a
small robot icon on the tablet once the child has finished correcting. In our
experiments, both were easy to grasp for children.

@@

Then, the actual implementation on the robot requires the coordination of
several modules (from performing gestures and acquiring the user's input to
the high-level state machine), spread over several devices (the robot itself,
one laptop and up to four tactile tablets for certain studies we conducted). We
relied on ROS to ensure the synchronization and communication between these
modules.

@@

By changing the eigenvalues associated to these eigenvectors and regenerating
letters with the reverse procedure, it becomes then easy to generate new
letters, with distortions that are actually plausible handwriting errors: they
are actually \emph{exaggerations} of variance of writing styles that naturally
occurs amongst adult writers.  Figure~\ref{fig:sampleLetters} shows examples of
deformed ``g'' generated with such a technique.

@@

The same technique can also be applied to \emph{classify} demonstrations,
\emph{assess} their quality and \emph{learn} from them. Figure~\ref{fig:h} shows
for example 9 allographs of ``h'' written by a 6 years old child, along with the
mean letter from the dataset (reference letter). By projecting each of the
demonstrations onto the eigenspace of ``h'' (Figure~\ref{fig:h}b), we observe
that:

@@

\begin{figure}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.4\columnwidth]{h}
    }
    \subfigure[Same samples, normalized and projected in the eigenspace spanned from the
        3 first eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.4\columnwidth]{eigenspace-uniformization}
    }

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child demonstration in the eigenspace.
Figure~\ref{learning_6_demos}, at the end of the article, illustrates the
process with the complete learning cycle of the number ``6''.

@@

The overall architecture of the system (Figure~\ref{fig:archi}) is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API\footnote{The ROS stack for {\sc nao} is available at
\url{http://wiki.ros.org/nao_robot}.} and the Aldebaran-provided NaoQI API, one
to four Android tablets (the main tablet is used to print the robot's letter and
to acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation and high level control of the activity.

Since the system does not actually require any CPU-intensive process, the laptop
can be removed and the whole logic run on the robot. Due to the relative
difficulty to deploy and debug ROS nodes directly on the robot, the laptop
remains however convenient during the development phase and we kept if during
our experiments.

Most of the nodes are written in Python, and the whole source code of the
project is available online\footnote{The primary repository is
\url{https://github.com/chili-epfl/cowriter_letter_learning}.}

@@

The system has been deployed and tested in several situations: in three
different schools (resulting in relatively short child-robot interactions with
more than 70 children, aged 5 to 8), in the surgery of an occupational
therapist (with 8 children, each interacting several hours with the system),
and for two case studies that each lasted several weeks. Table~\ref{studies}
gives an overview of these studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}     & {\bf Type}                                    & {\bf Avg. duration}              & {\bf Children \#}                     & {\bf Ages} \\ \midrule
{\it ISG 1}     & Unstructured group interaction at school      & 16 min/group               & 4 $\times$ 8 children                 & 6-7        \\
{\it Florimont} & Individual/Pair interaction at school         & 11 min/group               & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it Vincent}     & Case-study, spanning over 4 weeks              & 1.5h $\times$ 4 weeks & 1                                     & 6          \\
{\it Haut-Lac}  & Pair interaction at school & 26 min/group                            & 7 $\times$ 2                          & 5-6        \\
{\it ISG 2}     & Individual interaction at school              & 18 min                            & 6                                     & 5-6        \\
{\it Thomas}     & Case-study, spanning over 4 weeks              & 1h $\times$ 4 weeks        & 1                                     & 5          \\
{\it Normandie}     & Individual interactions at surgery               & 1h $\times$ 3 sessions & 8                                     & 6-8          \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
    \centering
    \includegraphics[height=3.9cm]{schools}
    \includegraphics[height=3.9cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    above 70 pupils interacting with the robot over short periods (around 15
minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

@@

The experiment took place at the surgery of an occupational therapist in
Normandy, France. Eight children participated, all showing troubles with
handwriting acquisition: Valentine (7 years old), Alexandre (6.5) and Jonathan
(7) are under the care of the occupational therapist. Enzo (8) and Matenzo (7)
are repeating their school year because of writing difficulties. Mona (6) and
Adele (8) are both ranked at the bottom of their respective classes in writing
activities. Nathan (7) is under the care of a neurologist, and has been
diagnosed with specific language impairment. Given their school level, all those
children would be expected to know how to correctly shape cursive letters.

@@

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Child & demo & rew & pun & p(robot) & p(child)\\ \hline
        Valentine & 127 & 24 & 6 & 2.4e-03 & 5.5e-02\\ \hline
        Enzo & 223 & 20 & 9 & 1.7e-01 & 3.5e-01\\ \hline
        Matenzo & 131 & 10 & 3 & 3.8e-03 & 7.9e-03\\ \hline
        Jonathan & 98 & 10 & 5 &  1.5e-01 & 3.8e-01\\ \hline
        Nathan & 115 & 16 & 4 & 5.3e-04 & 2.7e-03\\ \hline
        Alexandre & 83 & 10 & 3 & 3.1e-02 & 6.0e-01\\ \hline
        Adele & 35 & 4 & 2 & 5.0e-02 & 3.7e-02\\ \hline
        Mona & 40 & 5 & 1 &  5.4e-01 & 2.0e-01\\ \hline
    \end{tabular}
    \caption{\small Results of evaluations. demo: number of demonstrations provided by
        the child over all session. rew: number of rewards accorded by
        the child. pun: number of punishments. p(robot): are the evaluations
        significantly corresponding to the progression of the robot ? p(child): are the
        evaluations significantly corresponding to the child's own
progression ?}
    \label{table:scores}
\end{table}

@@

During this first case study, we invited and followed Vincent, a six years old
child, once per week over a period of a month. Our primary aim was to address the
question of whether we can sustain Vincent's engagement and commitment to the
writing activity over such a period.

@@

The study took place at our laboratory (Figure~\ref{fig:diego}), and we chose to
design the activity around a storyline meant to be attractive for a 6 years old
boy: one of our {\sc nao} was away for a mysterious scientific mission, and it
needed the support of another one -- which stayed at the lab -- to interpret
curious pictures that were sent every week. Vincent had to help the second robot
understanding the pictures, and since the two robots had somehow beforehand
agreed to communicate with ``letters written like humans'' (\ie handwritten),
Vincent also had to help the robot to write good looking letters (because, well,
this robot was terrible at writing!) The experimental setup was similar to
Figure~\ref{experimental_setup}, except that Vincent had to tell the robot what
to write with small plastic letters (visible behind the robot on Figure~\ref{fig:diego}).

@@

\begin{figure}
    \centering
    \subfigure[Initial letter, generated by the robot]{
        \includegraphics[height=6cm]{diego-initial-letter}
    }
    \subfigure[Final letter, after training with Vincent]{
        \includegraphics[height=6cm]{diego-final-letter}
    }

    \caption{\small (French) text generated by the robot, before and after a one
        hour long interaction session with the child. As an example, the red box
        highlights the changes on the word ``envoyer''.}

@@

Overall, Vincent provided 154 demonstrations to the robot, and he remained
actively engaged over the four weeks. The story was well accepted by Vincent and
he seriously engaged into the game. After the first week, he showed good
confidence to play with the robot and he built affective bonds with the robot
over the course of the study, as evidenced by some cries on the last session,
and several letters sent to the robot \emph{after} the end of the study
(one of them 4 months later) to get news. This represents a promising initial
result: we can effectively keep a child engaged with the robot for a relatively
long periods of time (about 5 hours).

No conclusion can be drawn in terms of actual handwriting remediation: we did
not design this study to formally assess possible improvements. However, as pictured on Figure~\ref{fig:stimuli}, Vincent was able to
significantly improve the robot's skill, and he acknowledged that he had been
able to help the robot: in that regard, Vincent convinced himself that he was
``good enough'' at writing to help someone else, which is likely to have
a positive impact on his self-esteem.

@@

Thomas is a 5.5 years old child, under the care of an occupational therapist. He
has been diagnosed with visuo-constructive deficits, which translate into
difficulties for him to consistently draw letters. Besides, focusing on a task
is difficult for Thomas, who tends to rapidly shift his attention to something
else. Since the robot's learning algorithm requires repeated demonstrations of
similarly shaped letters to converge, the occupational therapist was especially
interested in observing if the robot would induce a strong enough motivation for
Thomas to focus on producing regular, consistent letters, thus overcoming his
deficit.

The experiment took place at the therapist's surgery (four sessions spanning
over 5 weeks). Contrary to Vincent's experiment, we chose not to introduce any
storyline beyond a simple prompt (``the robot wants to participate to a robotic
handwriting contest, will you help him to prepare?''), only provided during the
first session. Hence, we also tested during this case-study if the robot (and
the Protégé effect) would induce a strong enough intrinsic motivation to
keep the child engage over the five weeks.

@@

Since Thomas would frequently draw mirrored numbers, or hard-to-recognize
shapes, the learning algorithm of the robot tended to converge towards
meaningless scribblings. We addressed this issue by having the robot to
\emph{refuse} allographs that were too far from the reference shape (\emph{``I'm
not sure to understand what you are drawing...''}), so that the child had to
pay good attention to what he would demonstrate to the robot. Also, to make
the robot's progresses evident, we modified the initialization step of the
learning algorithm to start with a roughly vertical stroke instead of a
deformed number (round 0 on Figure~\ref{learning_6_demos}).

@@

As soon as Thomas understood that the robot was only accepting well-formed
allographs, he started to focus on it and he would typically draw 5 or 6 times
the number before actually sending to the robot (the tablet lets children
clear their drawing and try again before sending it). According to
the therapist, it was the first time that Thomas would correct himself in such a
way, explicitly having to reflect on how \emph{another agent} (the robot) would
interpret and understand his writing. Figure~\ref{Thomas_progress} shows how
he gradually improved his demonstrations for some numbers, according to the
metric we used to make the robot accept/refuse samples.

Since the robot's handwriting started from a simple primitive (a stroke), each
time Thomas succeeded to have his demonstration accepted by it, the robot's
improvement was clearly visible (as measured in Figure~\ref{Thomas_distances}).
This led to a self-rewarding situation that effectively supported Thomas'
engagement.

@@

\begin{figure}
    \centering
    \subfigure[Number ``2'']{
        \includegraphics[width=0.45\linewidth]{henry2}
    }
	\subfigure[Number ``5'']{
        \includegraphics[width=0.45\linewidth]{henry5}
    }
    \caption{\small Improvement of Thomas demonstrations for some numbers: a) drawing the allograph of the number 1, b) the number 2 and c) the number 5. Thomas progressively took care of the demonstrations he was providing to the robot for those numbers. We used for this figure the same metric than the one used for the acceptance algorithm. Distances are normalized with respect to the biggest value. The dashed line correspond to the threshold of robot's acceptance.}

>>>>

Thomas is five and half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
workaround his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{Children's names have been changed.}

While Thomas is lively and always quick at shifting his attention from one
activity to another, Vincent is shy and poised. Two very different children,
facing however the same difficulty to write in a legible manner. And, hidden
beyond this impaired skill, psycho-social difficulties arise: they
underperform at school, Thomas has to go for follow-up visits every week,
they both live under the label ``requires special care''. This is a source of
anxiety for the children and for their parents alike.

@@

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks~\cite{Hoy2011}), essentially consisting in
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving his/her handwriting skills and,
consequently, improving his/her performance at school.

We present in this article a new take on this educative challenge: a remediation
procedure that involve a ``bad writer'' robot that is \emph{taught} by the
child. By building on the \emph{learning by teaching} paradigm, not only the
child practises handwriting but, as (s)he takes on the role of the teacher,
(s)he also positively reinforce his/her self-esteem and motivation: his/her
social role shifts from the ``underperformer'' to ``the one who knows and
teaches''. And by relying on a robot, we can tailor the exercises and the
learning rate to each children' needs, as we show in this article.

@@

We also investigate here a particular role for a robot in the
education of handwriting: not only is the robot actively performing the activity
by drawing letters, but it does so in a way that engages the child in a very
specific social role. The child is the teacher in this relationship and the
robot is the learner: the child must engage in a (meta-) cognitive relationship
with the robot to try to understand why the robot fails and how to help it best.
Here, the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphising, and
cognitively engage the child into the learning activity (be it consciously or
not).

Besides the commitment of the child into the interaction build on a
psychological effect known as the
``protégé effect''~\cite{Chase2009}: the teacher feels responsible for his student, commits
to the student's success and possibly experiences student's failure as his own
failure to teach. Teachable computer-based agents have previously been used to
encourage this ``protégé effect'', wherein students invest more effort into
learning when it is for a teachable agent than for themselves~\cite{Chase2009}.
We rely on this cognitive mechanism to reinforce the child's commitment into the
robot-mediated handwriting activity, and we indeed show sustained child-robot
engagement over extended periods of time (several hours spread of a month).

@@

In our case, the role of agency is stronger: it induces meta-cognition (``I am
interacting with an agent, so I need to reflect on how to best teach him'')
which is beneficial for the learning process; it also induces a protégé
effect (``I want my robot-agent to succeed!'') which support the commitment of
child into the interaction, also for longer period of time.

@@

The following sections of the paper go into details. We first provides a
brief overview of the robotic system and the interaction it induces. We also
present the machine-learning techniques that allow the robot to learn
from the children, followed by the implementation on a {\sc nao} robot.

We then present and report on the field experiments that we conducted over the
last two years, including four studies at school, one longer experiment
with eight children at the surgery of an occupational therapist, and two
one-month long case studies. While the focus of the school
experiments was mostly technical validation and data acquisition, the three
other experiments involved children with actual deficits, and gave us initial insights on the
relevance and effectiveness of our approach.

@@

Figure~\ref{experimental_setup} illustrates our general experimental setup: a
face-to-face child-robot interaction with an autonomous Aldebran's {\sc nao}
robot.

A tactile tablet (with a custom application) is used for both the robot and the
child to write: during a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and push the tablet
towards the robot. The robot writes on the tablet by gesturing the writing -- the
letters being actually drawn by the tablet application itself. The child then pull back the
tablet, corrects the robot's attempt by writing him/herself on top or next to
the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
demonstration to the robot by pressing a small button on the tablet. The robot
``learns'' from this demonstration and tries again.

Since the children are assumed to take on the role of the teachers, we had to ensure
they would be able to manage by themselves the turn-taking and the overall
progression of the activity (moving forwards to the next letter or word). In our design,
the turn-taking relies on the robot prompting for feedback once it is done with
its writing (through simple sentences like ``What do you think?''), and pressing on a
small robot icon on the tablet once the child has finished correcting. In our
experiments, both were easy to grasp for children.

@@

Then, the actual implementation on the robot requires the coordination of
several modules (from performing gestures and acquiring the user's input to
the state machine implementing the high-level behaviour), spread over several devices (the robot itself,
one laptop and up to four tactile tablets for some of the studies that we conducted). We
relied on ROS to ensure the synchronization and communication between these
modules.

@@

By changing the eigenvalues associated to these eigenvectors and regenerating
letters with the reverse procedure, it becomes then easy to generate new
letters, with distortions that are actually plausible handwriting errors: they
are actually exaggerations of variance of writing styles that naturally
occurs amongst adult writers.  Figure~\ref{fig:sampleLetters} shows examples of
deformed ``g'' generated with such a technique.

@@

The same technique can also be applied to \emph{classify} demonstrations,
\emph{assess} their quality and \emph{learn} from them. Figure~\ref{fig:h} shows
for example nine allographs of ``h'' written by a 6 years old child, along with the
mean letter from the dataset (reference letter). By projecting each of the
demonstrations onto the eigenspace of ``h'' (Figure~\ref{fig:h}b), we observe
that:

@@

\begin{figure}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.45\columnwidth]{h}
    }
    \subfigure[Same samples, normalized and projected in the eigenspace spanned from the
        3 first eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.45\columnwidth]{eigenspace-uniformization}
    }

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child demonstration in the eigenspace.
Figure~\ref{learning_6_demos} (p.~\pageref{learning_6_demos}) illustrates the
process with the complete learning cycle of the number ``6''.

@@

The overall architecture of the system (Figure~\ref{fig:archi}) is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API\footnote{The ROS documentation for {\sc nao} is available at
\url{http://wiki.ros.org/nao}.} and the Aldebaran-provided NaoQI API, one
to four Android tablets (the main tablet is used to draw the robot's letter and
to acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation (based on the NaoQI inverse kinematics
library) and high level control of the activity.

Since the system does not actually require any CPU-intensive process, the laptop
can be removed and the whole logic run on the robot. Due to the relative
difficulty to deploy and debug ROS nodes directly on the robot, the laptop
remains however convenient during the development phase and we kept it during
our experiments.

Most of the nodes are written in Python, and the whole source code of the
project is available online\footnote{The primary repository is
\url{https://github.com/chili-epfl/cowriter_letter_learning}.}. Further details
on the technical implementation are available in~\cite{hood2015when}.

@@

The system has been deployed and tested in several situations: in three
different schools (relatively short child-robot interactions with
more than 70 children, aged 5 to 8), in the surgery of an occupational
therapist (with 8 children, each interacting several hours with the system),
and in two case studies that each lasted several weeks. Table~\ref{studies}
gives an overview of these studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf Children \#}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 18 min                 & 6                                     & 5-6        \\ \midrule
{\it Surgery}      & Individual interaction at surgery        & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study, spanning over 4 weeks        & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study, spanning over 4 weeks        & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

We report hereafter the main design choices and results for each of these
studies and experiments. The interested reader can find supplementary details
in~\cite{jacq2016building, hood2015when}.

@@

\begin{figure}
    \centering
    \includegraphics[height=3.9cm]{schools}
    \includegraphics[height=3.9cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    more than 70 pupils interacting with the robot over short periods (around 15
minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

@@

The experiment took place at the surgery of an occupational therapist in
Normandy, France. Eight children participated, all showing troubles with
handwriting acquisition: Valérie (7 years old), Antoine (6.5) and Johan
(7) are under the care of the occupational therapist. Émilien (8) and Mathieu (7)
are repeating their school year because of writing difficulties. Mona (6) and
Adèle (8) are both ranked at the bottom of their respective classes in writing
activities. Nicolas (7) is under the care of a neurologist, and has been
diagnosed with specific language impairment. Given their school level, all those
children would be expected to know how to correctly shape cursive letters.

@@

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        child & demo & rew & pun & p(robot) & p(child)\\ \hline
        Valérie & 127 & 24 & 6 & 2.4e-03 & 5.5e-02\\ \hline
        Émilien & 223 & 20 & 9 & 1.7e-01 & 3.5e-01\\ \hline
        Mathieu & 131 & 10 & 3 & 3.8e-03 & 7.9e-03\\ \hline
        Johan & 98 & 10 & 5 &  1.5e-01 & 3.8e-01\\ \hline
        Nicolas & 115 & 16 & 4 & 5.3e-04 & 2.7e-03\\ \hline
        Antoine & 83 & 10 & 3 & 3.1e-02 & 6.0e-01\\ \hline
        Adèle & 35 & 4 & 2 & 5.0e-02 & 3.7e-02\\ \hline
        Marie & 40 & 5 & 1 &  5.4e-01 & 2.0e-01\\ \hline
    \end{tabular}
    \caption{\small results of evaluations. demo: number of demonstrations provided by
        the child over all session. rew: number of rewards accorded by
        the child. pun: number of punishments. p(robot): are the evaluations
        significantly corresponding to the progression of the robot ? p(child): are the
        evaluations significantly corresponding to the child's own
progression ?}
    \label{table:scores}
\end{table}

@@

For the first case study, we invited and followed Vincent, a six years old
child, once per week over a period of a month. Our primary aim was to address the
question of whether we can sustain Vincent's engagement and commitment to the
writing activity over such a period.

@@

The study took place at our laboratory (Figure~\ref{fig:diego}), and we chose to
design the activity around a storyline meant to be attractive for a 6 years old
boy: one of our {\sc nao} was away for a mysterious scientific mission, and it
needed the support of another one -- which would remain at the lab -- to interpret
curious pictures that were sent every week. Vincent had to help the second robot
understanding the pictures, and since the two robots had somehow beforehand
agreed to communicate with ``letters written like humans'' (\ie handwritten),
Vincent also had to help the robot to write good looking letters (because, well,
this robot was terrible at writing!) The experimental setup was similar to
Figure~\ref{experimental_setup}, except that Vincent had to tell the robot what
to write with small plastic letters (visible behind the robot on Figure~\ref{fig:diego}).

@@

\begin{figure}
    \centering
    \subfigure[Initial text, generated by the robot]{
        \includegraphics[height=6cm]{diego-initial-letter}
    }
    \subfigure[Final text, after training with Vincent]{
        \includegraphics[height=6cm]{diego-final-letter}
    }

    \caption{\small (French) text generated by the robot, before and after a one
        hour long interaction session with the child. The red box
        highlights one instance of striking improvement of the robot's
        handwriting legibility.}

@@

Over the whole duration of the study, Vincent had provided 154 demonstrations to the robot, and he remained
actively engaged over the four weeks. The story was well accepted by the child and
he seriously engaged into the game. After the first week, he showed good
confidence to play with the robot and by the end of the study he had built
affective bonds with the robot, as evidenced by several letters he did send to the
robot \emph{after} the end of the study (one of them four months later) to get
news. This represents an initial validation of our hypothesis: our system can effectively keep a
child engaged with the robot for a relatively long periods of time (about 5
hours spread over a month).

No hard conclusion can be drawn in terms of actual handwriting remediation: we did
not design this study to formally assess possible improvements. However, as
visible on Figure~\ref{fig:stimuli}, Vincent was able to significantly improve
the robot's skill, and post-hoc interviews revealed that he acknowledged that he
was the one helping the robot. In that respect, Vincent realized that he was
``good enough'' at writing to help someone else. Quoting the mother a week
after the end of the experiment: ``Vincent's handwriting has changed over the
last weeks, going from a mix of standalone and cursive letters to full words in
cursive. This requires a lot of efforts and concentration from him, but he did
succeed during the sessions with the robot as he knew he had to show a
consistent writing to the robot''.

@@

Thomas is a 5.5 years old child, under the care of this occupational therapist. He
has been diagnosed with visuo-constructive deficits, which translate into
difficulties for him to consistently draw letters. Besides, focusing on a task
is difficult for Thomas, who tends to rapidly shift his attention to other
things. Since the robot's learning algorithm requires repeated demonstrations of
similarly shaped letters to converge, the occupational therapist was especially
interested in observing if the robot would induce a strong enough motivation for
Thomas to focus on producing regular, consistent letters, thus overcoming his
deficit.

The experiment took place at the therapist's surgery (four sessions spanning
over 5 weeks). Contrary to Vincent's experiment, we chose not to introduce any
storyline beyond a simple prompt (``the robot wants to participate to a robotic
handwriting contest, will you help him to prepare?''), only provided during the
first session. Hence, we also tested during this case-study if the robot (and
the Protégé effect) would induce by itself a strong enough intrinsic motivation to
keep the child engaged over the five weeks.

@@

Since Thomas would frequently draw mirrored numbers, or hard-to-recognize
shapes, the learning algorithm of the robot initially tended to converge towards
meaningless scribblings. We addressed this issue by having the robot to
\emph{refuse} allographs that were too far from the reference shape (the robt
would instead say \emph{``I'm
not sure to understand what you are drawing...''}), so that the child had to
pay good attention to what he would demonstrate to the robot. Also, to make
the robot's progresses evident, we modified the initialization step of the
learning algorithm to start with a roughly vertical stroke instead of a
deformed number (round 0 on Figure~\ref{learning_6_demos}).

@@

As soon as Thomas understood that the robot was only accepting well-formed
allographs, he started to focus on it and he would typically draw 5 or 6 times
the number before actually sending to the robot (the tablet lets children
clear their drawing and try again before sending it). According to
the therapist, it was the first time that Thomas would correct himself in such a
way, explicitly having to reflect on how \emph{another agent} (the robot) would
interpret and understand his writing. Figure~\ref{Thomas_progress} shows how
he gradually improved his demonstrations for two different numbers, according to the
metric we used to make the robot accept/refuse samples.

Since the robot's handwriting started from a simple primitive (a stroke), each
time Thomas succeeded to have his demonstration accepted by it, the robot's
improvement was clearly visible (as shown in Figure~\ref{Thomas_distances}).
This led to a self-rewarding situation that effectively supported Thomas'
engagement.

@@

\begin{figure}
    \centering
    \subfigure[Number ``2'']{
        \includegraphics[width=0.45\linewidth]{henry2}
    }
	\subfigure[Number ``5'']{
        \includegraphics[width=0.45\linewidth]{henry5}
    }
    \caption{\small Improvement of Thomas demonstrations for the numbers ``2''
        and ``5''. Thomas progressively took care of the demonstrations he was providing to the
    robot for those numbers. We used for this figure the same metric than the one
    used for the acceptance algorithm. Distances are normalized with respect to the
    biggest value. The dashed line correspond to the threshold of robot's
    acceptance.}
